On the uniqueness and stability of dictionaries \\ for sparse representation of noisy signals

### 2019-05-08 manuscripts/IEEE/final_submission.tex
discussion edits

One direct application of our work is to neural theory, wherein our theorems justify the mathematical soundness of one of the few hypothesized theories of bottleneck communication in the brain \cite{Isely10}: that sparse neural population activity is recoverable from its noisy linear compression through a randomly constructed (but unknown) wiring bottleneck by any biologically plausible unsupervised sparse coding method that solves Prob.~\ref{DeterministicUniquenessTheorem} or \ref{SLCopt} (e.g., \cite{rehnsommer2007, rozell2007neurally, pehlevan2015normative}).\footnote{We refer the reader to \cite{ganguli2012compressed} for more on interactions between dictionary learning and neuroscience.}

@@

Beyond an extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime and their novel application to Prob.~\ref{OptimizationProblem}, our work contains a theory of combinatorial designs for support sets key to the identification of the dictionary. We incorporate this idea into a fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) that draws upon the definition of a new matrix lower bound induced by a hypergraph. New insights enabled by this combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) recoverable dictionaries need not necessarily satisfy the spark condition.

The absence of any assumption at all about dictionaries that solve Prob.~\ref{InverseProblem} was a major technical difficulty in proving Thm.~\ref{DeterministicUniquenessTheorem}. We sought such a general guarantee because of the practical difficulty of ensuring that an algorithm maintain a dictionary satisfying the spark condition \eqref{SparkCondition} at each iteration, an implicit requirement of all previous works except \cite{Hillar15}; indeed, even certifying a dictionary has this property is NP-hard \cite{tillmann2014computational}.

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks; e.g., for MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and even the design of an ultrafast camera \cite{Gao14}. It is only a matter of time before these systems incorporate dictionary learning to encode and decode signals (e.g., in a device that learns structure from motion \cite{kong2016prior}), just as scientists have used it to make sense of their data \cite{jung2001imaging, Agarwal14, lee2016sparse, wu2016stability}.

Assurances such as those offered by our theorems certify that different devices and algorithms (with different initialization, etc.) will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these models of data; moreover, even small noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis seems to be the assumed well-posedness of parameter identification in the sparse coding model, confirmation of which forms the core of our theoretical findings.

>>>>

One direct application of this work is to theoretical neuroscience, wherein our theorems justify the mathematical soundness of one of the few hypothesized theories of bottleneck communication in the brain \cite{Isely10}: that sparse neural population activity is recoverable from its noisy linear compression through a randomly constructed (but unknown) wiring bottleneck by any biologically plausible unsupervised sparse coding method that solves Prob.~\ref{DeterministicUniquenessTheorem} or \ref{SLCopt} (e.g., \cite{rehnsommer2007, rozell2007neurally, pehlevan2015normative}).\footnote{We refer the reader to \cite{ganguli2012compressed} for more on interactions between dictionary learning and neuroscience.}

@@

Beyond an extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime and their novel application to Prob.~\ref{OptimizationProblem}, our work contains a theory of combinatorial designs for support sets key to identification of dictionaries. We incorporate this idea into a fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) that draws upon the definition of a matrix lower bound induced by a hypergraph. New insights enabled by this combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) recoverable dictionaries need not satisfy the spark condition.

The absence of any assumptions about dictionaries solving Prob.~\ref{InverseProblem} was a major technical difficulty in proving Thm.~\ref{DeterministicUniquenessTheorem}. We sought such a general guarantee because of the practical difficulty in ensuring that an algorithm maintain a dictionary satisfying the spark condition \eqref{SparkCondition} at each iteration, an implicit requirement of all previous works except \cite{Hillar15}; indeed, even certifying a dictionary has this property is NP-hard \cite{tillmann2014computational}.

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks; for instance, in MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and even the design of an ultrafast camera \cite{Gao14}. It is only a matter of time before these systems incorporate dictionary learning to encode and decode signals (e.g., in a device that learns structure from motion \cite{kong2016prior}), just as scientists have used sparse coding to make sense of their data \cite{jung2001imaging, agarwal2014spatially, lee2016sparse, wu2016stability}.

Assurances such as those offered by our theorems certify that different devices and algorithms will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these models of data; moreover, even small noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis seems to be the assumed well-posedness of parameter identification in the sparse coding model, confirmation of which forms the core of our theoretical findings.
